Sequential Composition for Relaxed Memory: Pomsets with Predicate Transformers

### 2021-07-16 paper/oopsla.tex
cleanup discussion of PS

Recently, \citet{promising-ldrf} showed that certain combinations of compiler
optimizations are inconsistent with local \drf{} guarantees.  All of the
examples that prove inconsistency have the same shape: they combine read
introduction and case analysis (aka, if-closure).  Effectively, this turns
one read into two, where different conditional branches can be taken for the
two copies of the read.  This is reminiscent of the type of \emph{bait and
  switch} behavior noted by \citet{DBLP:journals/pacmpl/JagadeesanJR20}: the
promising semantics (\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related
models
\citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
fail to validate compositional reasoning of temporal properties.  Consider
example \ref{oota4} from \cite{DBLP:journals/pacmpl/JagadeesanJR20}:
\begin{align}
  \taglabel{oota4}
  \begin{gathered}
    \PW{y}{x}
    \PAR
    \PR{y}{r} \SEMI \PR{b}{s} \SEMI\IF{s}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
    \PAR
    \PW{b}{1}
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{1}}{right=of rx}
        \po{rx}{wy}
        \event{ry}{\DR{y}{1}}{right=3em of wy} 
        \event{wx}{\DW{x}{1}}{right=of ry}
        \event{wz}{\DW{z}{1}}{right=of wx}
        \event{rb}{\DR{b}{1}}{right=of wz}
        \event{wb1}{\DW{b}{1}}{right=3em of rb}
        \po{ry}{wx}
        \rf{wb1}{rb}
        \rf{wy}{ry}
        \rf[out=-170,in=-10]{wx}{rx}
        \po{rb}{wz}
        \po[out=15,in=165]{ry}{wz}
      \end{tikzinline}}
  \end{gathered}  
\end{align}
Under \PS{}, this outcome is allowed by \emph{baiting} with the \texttt{else}
branch, then \emph{switching} to the \texttt{then} branch, based on a coin
flip $(\IF{b})$.
% Using a combination of read introduction and case analysis,
% the middle thread can be rewritten to:
% \begin{displaymath}
%   \begin{array}{l}
%     \PR{y}{r} \SEMI \IF{b}\THEN\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \ELSE\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \FI
%   \end{array}
% \end{displaymath}
% \footnote{Call the threads \texttt{s}, \texttt{t}, and
%   \texttt{u}.  To get the result in the promising semantics, first execute
%   \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
%   \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
%   Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
%   \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
%   fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
%   the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
%   removing timestamps and replacing the word \emph{promise} by
%   \emph{speculation}.}
% \citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
% of temporal reasoning could result in violations of the Java security
% architecture.

Under all variants of \PwT{}, this outcome is disallowed, due to a cycle in
${\rrfx}\cup{\lt}$. (Note that all reads cross threads in the example, so
there is no difference between \PwTmca{1} and \PwTmca{2}.)

The examples in \cite{promising-ldrf} are more complex.  Here is one, dubbed
\textsc{ldrf-fail-ps}:
\begin{gather*}  
  %\taglabel{LDRF-Fail-PS}
  \begin{gathered}
  \IF{x}\THEN
    \PFADD{w}{}{1}\SEMI
    \PW{y}{1}\SEMI
    \PW{z}{1}
  \FI
  \PAR
  \IF{\BANG z}\THEN
    \PW{x}{1}
  \ELSE
    \IF{\BANG \PFADD{w}{}{1}}\THEN
      \PW{x}{\PR{y}{}}
    \FI
  \FI
    \\
    \hbox{\begin{tikzinlinesmall}[node distance=2em]
        \event{a1}{\DR{x}{1}}{}
        \event{a2}{\DR{w}{1}}{right=of a1}
        \event{a3}{\DW{w}{2}}{right=of a2}
        \event{a4}{\DW{y}{1}}{right=of a3}
        \event{a5}{\DW{z}{1}}{right=of a4}
        \event{b1}{\DR{z}{1}}{right=5em of a5}
        \event{b2}{\DR{w}{0}}{right=of b1}
        \event{b3}{\DW{w}{1}}{right=of b2}
        \event{b4}{\DR{y}{1}}{right=of b3}
        \event{b5}{\DW{x}{1}}{right=of b4}
        \rmw{a2}{a3}
        \po[out=15,in=165]{a1}{a3}
        \po[out=15,in=165]{a1}{a4}
        \po[out=15,in=165]{a1}{a5}        
        \rmw{b2}{b3}
        \po{b4}{b5}
        \po[out=15,in=165]{b2}{b5}        
        \po[out=15,in=165]{b1}{b3}
        \rf{a5}{b1}
        \rf[out=15,in=165]{a4}{b4}
        \rf[out=-165,in=-15]{b3}{a2}
        \rf[out=-165,in=-15]{b5}{a1}
      \end{tikzinlinesmall}}
  \end{gathered}
\end{gather*}
This execution is allowed by \PS{} and disallowed by \PwTmca{}.

The difference between \PS{} and \PwT{} can be understood in terms of the
valid program transformations.  \PS{} allows reads to be introduced, with
subsequent case analysis on the value read.  \PwT{} validates case analysis,
but invalidates read introduction.  In return, \PwT{} enjoys compositionality
for temporal safety properties, as discussed in \textsection\ref{sec:results}.

\citet{promising-ldrf} show that by restricting \RMW{}-store reorderings, one
can establish local \drf{} for \PS{}.  We believe that no such restriction is
required for \PwT{}.

These differences highlight the subtle tensions between compiler optimizations
and program logics that are revealed by relaxed memory models.  It is not
possible to have everything one wants. Thus, one is forced to choose which
optimizations and reasoning principles are most important.

@@

% Case analysis gives very weak results when combined with thread inlining.
% See \cite[\textsection B.1]{DBLP:journals/pacmpl/ChakrabortyV19appendix}.
% These happen by performing transformations that: 
% (1) introduce conditionals,
% (2) inline two threads on both sides of the introduced conditional,
% (3) choose different orders for the two threads for the two sides of the conditional.

@@

>>>>

Recently, \citet{promising-ldrf} showed that certain combinations of compiler
optimizations are inconsistent with local \drf{} guarantees.  All of the
examples that prove inconsistency have the same shape: they combine read
introduction and case analysis (aka, if-closure).  Effectively, this turns
one read into two, where different conditional branches can be taken for the
two copies of the read.  This is reminiscent of the type of \emph{bait and
  switch} behavior noted by \citet{DBLP:journals/pacmpl/JagadeesanJR20}: the
promising semantics (\PS{}) \cite{DBLP:conf/popl/KangHLVD17} and related
models
\citep{DBLP:conf/esop/JagadeesanPR10,DBLP:journals/pacmpl/ChakrabortyV19,Manson:2005:JMM:1047659.1040336},
fail to validate compositional reasoning of temporal properties.  Consider
example \ref{oota4} from \cite{DBLP:journals/pacmpl/JagadeesanJR20}:
\begin{gather*}
  \taglabel{oota4}
  \begin{gathered}
    \PW{y}{x}
    \PAR
    \PR{y}{r} \SEMI\IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI
    \PAR
    \PW{b}{1}
    % \\[-1ex]
    % \hbox{\begin{tikzinline}[node distance=1.5em]
    %     \event{rx}{\DR{x}{1}}{}
    %     \event{wy}{\DW{y}{1}}{right=of rx}
    %     \po{rx}{wy}
    %     \event{ry}{\DR{y}{1}}{right=3em of wy} 
    %     \event{wx}{\DW{x}{1}}{right=of ry}
    %     \event{wz}{\DW{z}{1}}{right=of wx}
    %     \event{rb}{\DR{b}{1}}{right=of wz}
    %     \event{wb1}{\DW{b}{1}}{right=3em of rb}
    %     \po{ry}{wx}
    %     \rf{wb1}{rb}
    %     \rf{wy}{ry}
    %     \rf[out=-170,in=-10]{wx}{rx}
    %     \po{rb}{wz}
    %     \po[out=15,in=165]{ry}{wz}
    %   \end{tikzinline}}
    \\[-1ex]
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{rx}{\DR{x}{1}}{}
        \event{wy}{\DW{y}{1}}{right=of rx}
        \po{rx}{wy}
        \event{rb}{\DR{b}{1}}{right=3em of wy}
        \event{ry}{\DR{y}{1}}{right=of rb} 
        \event{wx}{\DW{x}{1}}{right=of ry}
        \event{wz}{\DW{z}{1}}{right=of wx}
        \event{wb1}{\DW{b}{1}}{right=3em of wz}
        \po{ry}{wx}
        \rf[out=-170,in=-10]{wb1}{rb}
        \rf[out=15,in=165]{wy}{ry}
        \rf[out=-170,in=-10]{wx}{rx}
        \po[out=15,in=165]{rb}{wz}
        \po[out=15,in=165]{ry}{wz}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
Under all variants of \PwT{}, this outcome is disallowed, due to the cycle
involving $x$ and $y$.\footnote{All of the reads in \ref{oota4} are
  cross-thread, so there is no difference between \PwTmca{1} and \PwTmca{2}.
  For \PwTc, there is a cycle in ${\rrfx}\cup{\lt}$.}  Under \PS{}, this
outcome is allowed by {baiting} with the \texttt{else} branch, then
{switching} to the \texttt{then} branch, based on a coin flip (${b}$).
% Using a combination of read introduction and case analysis,
% the middle thread can be rewritten to:
% \begin{displaymath}
%   \begin{array}{l}
%     \PR{y}{r} \SEMI \IF{b}\THEN\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \ELSE\\
%     \quad \IF{b}\THEN  \PW{x}{r} \SEMI \PW{z}{r} \ELSE \PW{x}{1} \FI\\
%     \FI
%   \end{array}
% \end{displaymath}
% \footnote{Call the threads \texttt{s}, \texttt{t}, and
%   \texttt{u}.  To get the result in the promising semantics, first execute
%   \texttt{u} to get message \texttt{<b:1@1>}.  Then \texttt{t} promises
%   \texttt{<x:1@1>}, which it can fulfill by reading \texttt{b}$=$\texttt{0}.
%   Then execute \texttt{s} to get message \texttt{<y:1@1>}.  Then execute
%   \texttt{t}, reading \texttt{b}$=$\texttt{1} and \texttt{y}$=$\texttt{1} and
%   fulfill the promise by writing \texttt{<x:1@1>}. The execution is exactly
%   the same in our speculative semantics \cite{DBLP:conf/esop/JagadeesanPR10},
%   removing timestamps and replacing the word \emph{promise} by
%   \emph{speculation}.}
% \citet[Fig.~8]{DBLP:journals/toplas/Lochbihler13} notes that such violations
% of temporal reasoning could result in violations of the Java security
% architecture.

\citet{promising-ldrf} introduce more complex examples to show that the
promising semantics fails \ldrfsc{}.\footnote{\citet{promising-ldrf} show
  that by restricting \RMW{}-store reorderings, one can establish \ldrfsc{}
  for \PS{}.  We speculate that no such restriction is required for \PwT{}.
  (We did not treat \RMW{}s in our proof of \ldrfsc{}.)}  Here is one, dubbed
\labeltext{\textsc{ldrf-fail-ps}}{LDRF-Fail-PS}.
\begin{gather*}  
  %\taglabel{LDRF-Fail-PS}
  \begin{gathered}
  \IF{x}\THEN
    \PFADD{w}{}{1}\SEMI
    \PW{y}{1}\SEMI
    \PW{z}{1}
  \FI
  \PAR
  \IF{z}\THEN
    \IF{\BANG \PFADD{w}{}{1}}\THEN
      \PW{x}{\PR{y}{}}
    \FI
  \ELSE
    \PW{x}{1}
  \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1.5em]
        \event{a1}{\DR{x}{1}}{}
        \event{a2}{\DR{w}{1}}{right=of a1}
        \event{a3}{\DW{w}{2}}{right=of a2}
        \event{a4}{\DW{y}{1}}{right=of a3}
        \event{a5}{\DW{z}{1}}{right=of a4}
        \event{b1}{\DR{z}{1}}{right=4em of a5}
        \event{b2}{\DR{w}{0}}{right=of b1}
        \event{b3}{\DW{w}{1}}{right=of b2}
        \event{b4}{\DR{y}{1}}{right=of b3}
        \event{b5}{\DW{x}{1}}{right=of b4}
        \po[out=15,in=165]{a1}{a3}
        \po[out=15,in=165]{a1}{a4}
        \po[out=15,in=165]{a1}{a5}        
        \po{b4}{b5}
        \po[out=15,in=165]{b2}{b5}        
        \po[out=15,in=165]{b1}{b3}
        \rf{a5}{b1}
        \rf[out=10,in=170]{a4}{b4}
        \rf[out=-170,in=-10]{b3}{a2}
        \rf[out=-170,in=-10]{b5}{a1}
        \rmw{a2}{a3}
        \rmw{b2}{b3}
      \end{tikzinline}}
  \end{gathered}
\end{gather*}
Again, all variants of \PwT{} disallow the outcome due to the cycle involving
$x$ and $y$.  It is allowed by \PS{} by baiting the second thread with
$\PW{x}{1}$ in the \texttt{else} branch, then switching to the \texttt{then}
branch.  This shows some some structural resemblance to \ref{oota4}, with $z$
replacing $b$.

\citeauthor{promising-ldrf} argue that the outcome of \ref{LDRF-Fail-PS} is
inevitable due to compiler optimizations.  The examples crucially involve the
following sequence of operations:
\begin{itemize}
\item read introduction,
\item if introduction, branching on the read just introduced.
\end{itemize}
We believe this combination of optimizations is unsound.  This is obviously
the case in \cXI: read introduction may cause undefined behavior (\ub), due to
the possible introduction of a data race.

The situation is more delicate in \llvm.  The short version of the story is
that load-hoisting followed by case analysis is unsound in \llvm, without freeze. 
This happens because:
\begin{itemize}
\item read introduction may result in the undefined value $\UNDEFINED$, due
  to the possible introduction of a data race \cite{DBLP:conf/cgo/ChakrabortyV17}, and
\item branching on an undefined value in \llvm{} results in \ub. 
\end{itemize}
\llvm{} delays \ub{} using the undefined value.  This allows \llvm{} to
perform optimizations such as load hoisting, where
\begin{math}
  \IF{C}\THEN \PR{x}{r}\FI
\end{math}
is rewritten to 
\begin{math}
  \PR{x}{s}\SEMI
  \LET{r}{\TERNARY{C}{s}{r}}.
\end{math}
Despite this, other optimizations regularly performed by \llvm{} are
unsound \cite{DBLP:conf/pldi/LeeKSHDMRL17}.  An example is loop switching,
where
\begin{math}
  \WHILE{C_1}\DO \IF{C_2}\THEN \aCmd_1 \ELSE \aCmd_2 \FI \OD
\end{math}
is rewritten to 
\begin{math}
  \IF{C_2}\THEN \WHILE{C_1}\DO \aCmd_1 \OD \ELSE \WHILE{C_1}\DO \aCmd_2 \FI \OD.
\end{math}
Freeze was introduced in \llvm{} in order to make such optimizations sound by
allowing branch on frozen $\UNDEFINED$ to give nondeterministic choice rather
than \ub{}.  In the RFC for freeze, \citet{nuno} says: \emph{``Note that
  having branch on poison not trigger \ub{} has its own problems.  We believe
  this is a good tradeoff.''}  \ref{LDRF-Fail-PS} demonstrates a concrete
problem with this tradeoff.  Other compilers, such as Compcert, are more
conservative \cite[\textsection9]{DBLP:conf/pldi/LeeKSHDMRL17}.

Thus, the difference between \PS{} and \PwT{} can be understood in terms of
the valid program transformations.  \PS{} allows reads to be introduced, with
subsequent case analysis on the value read.  \PwT{} validates case analysis,
but invalidates read introduction.

Allowing executions such as \ref{oota4} and \ref{LDRF-Fail-PS} also
invalidates compositional reasoning for temporal safety properties (see
\textsection\ref{sec:results}).

These differences highlight the subtle tensions between compiler
optimizations and program logics that are revealed by relaxed memory models.
It is not possible to have everything one wants. Thus, one is forced to
choose which optimizations and reasoning principles are most
important.\footnote{Another example is the tension between load
  hoisting---forbidden in \cXI{} but allowed by \llvm{}---and common
  subexpression elimination over an acquiring lock---allowed by \cXI{} but
  forbidden by \llvm{} \cite{DBLP:conf/cgo/ChakrabortyV17}.}

Finally, we note that it is possible that \PS{} is properly weaker than \PwT{}.

@@

% Case analysis gives very weak results when combined with thread inlining.
% See \cite[\textsection B.1]{DBLP:journals/pacmpl/ChakrabortyV19appendix}.
% These happen by performing transformations that: 
% (1) introduce conditionals,
% (2) inline two threads on both sides of the introduced conditional,
% (3) choose different orders for the two threads for the two sides of the ( conditional.

@@

% \subsubsection{Load hoisting in LLVM}
% Load-hoisting followed by case analysis is unsound in LLVM, without freeze.
% Introducing a read may cause a race, resulting in read value $\UNDEFINED$.
% Branch on $\UNDEFINED$ is UB.  Freeze was added to get around this...

% Examples from \cite{promising-ldrf} show that freeze is bullshit.  Compcert
% does not validate loop switching
% \cite[\textsection9]{DBLP:conf/pldi/LeeKSHDMRL17}.

% \cite[\textsection3.3]{DBLP:conf/pldi/LeeKSHDMRL17} Global Value Numbering
% (GVN) and Loop Unswitching require different semantics for branch on
% $\UNDEFINED$. \url{https://llvm.org/docs/LangRef.html#freeze-instruction}

% \href{https://lists.llvm.org/pipermail/llvm-dev/2016-October/106182.html}{Purpose of Freeze}
% \begin{quotation}
%   Poison is propagated aggressively throughout. However, there are cases
%   where this breaks certain optimizations, and therefore freeze is used to
%   selectively stop poison from being propagated.

%   A use of freeze is to enable speculative execution.  For example, loop
%   switching does the following transformation:
% \begin{verbatim}
% \end{verbatim}
% \begin{verbatim}
% while (C) {        if (C2) {     
%   if (C2) {           while (C)  
%    A                     A       
%   } else {   ==>   } else {      
%    B                   while (C) 
%   }                       B      
% }                  }             
% \end{verbatim}
%   Here we are evaluating C2 before C.  If the original loop never executed
%   then we had never evaluated C2, while now we do.  So we need to make sure
%   there's no UB for branching on C2.  Freeze ensures that so we would
%   actually have 'if (freeze(C2))' instead.  Note that
%   \emph{having branch on poison not trigger UB has its own problems.}
%   We believe this is a good tradeoff.
% \end{quotation}

% \begin{quotation}
%   LLVM frequently performs such load introductions in the “simplify CFG”
%   pass; e.g., when hoisting loads outside of conditionals.
% \end{quotation}

% case analysis happens in ``function specialization'' also known as
% ``procedure cloning''

% \subsubsection{Hoisting and CSE}
% Example from Viktor's talk: ``Weak Memory Concurrency in C/C++11 and LLVM''

% Load Hoisting:
% \begin{displaymath}
%   \IF{c}\THEN \PR{x}{a}\FI
%   \rightsquigarrow
%   \PR{x}{t}\SEMI
%   \LET{a}{\TERNARY{c}{t}{a}}
% \end{displaymath}

% CSE over acquiring lock:
% \begin{displaymath}
%   \PR{x}{a}\SEMI
%   \LOCK\SEMI
%   \PR{x}{b}
%   \rightsquigarrow
%   \PR{x}{a}\SEMI
%   \LOCK\SEMI
%   \LET{b}{a}
% \end{displaymath}

% Having both is clearly wrong:
% \begin{displaymath}
%   \begin{array}{l}
%     \IF{c}\THEN\\
%     \quad\PR{x}{a}\FI\SEMI\\
%     \LOCK\SEMI\\
%     \PR{x}{b}    
%   \end{array}
%   \rightsquigarrow
%   \begin{array}{l}
%     \PR{x}{t}\SEMI\\
%     \LET{a}{\TERNARY{c}{t}{a}}\SEMI\\
%     \LOCK\SEMI\\
%     \PR{x}{b}    
%   \end{array}
%   \rightsquigarrow
%   \begin{array}{l}
%     \PR{x}{t}\SEMI\\
%     \LET{a}{\TERNARY{c}{t}{a}}\SEMI\\
%     \LOCK\SEMI\\
%     \LET{b}{a}    
%   \end{array}
% \end{displaymath}
% When c is false, x is moved out of the critical region!

% So we have to forbid one transformation.
% \begin{itemize}
% \item C11 forbids load hoisting, allows CSE over lock().
% \item LLVM allows load hoisting, forbids CSE over lock().
% \end{itemize}
